Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-09-08 summary_minh_edited.tex
changes based on discussion from earlier

\author{John Koo\\Department of Statistics, Indiana University\\
Minh Tang\\Department of Statistics, North Carolina State University\\
Michael Trosset\\Department of Statistics, Indiana University}

@@

Let $G = (V, E)$ be an unweighted, undirected, 
and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. The matrix
$A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ 
such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ 
and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. 
We further restrict our analyses to Bernoulli graphs. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. 
Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ 
for each $1 \leq i < j \leq n$ (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). 
We denote $A \sim \BG(P)$ as a graph with adjacency matrix $A$ 
sampled from edge probability matrix $P$ in this manner. 
If each vertex has a hidden label in $[K]$, 
they are denoted as $z_1, ..., z_n$. 
Finally, we denote 
$X = \bigl[x_1 \mid \cdots \mid x_n\bigr]^\top \in \mathbb{R}^{n \times d}$ 
as the matrix corresponding to a collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

@@

Nevertheless by using the special geometric structure of $X$ we can circumvent the
non-identifiability of $Y$ and $XU$ by using instead the rows of the
matrix $V$ of eigenvectors (corresponding to the non-zero eigenvalues) of $P$. In particular $V$ is identifiable
up to orthogonal transformations and furthermore, due to the block
diagonal structure of $X$, the rows of $V$ also lie on $K$ distinct orthogonal
subspaces and hence $v_i^{\top} v_j = 0$ whenever $z_i \not = z_j$. 
% The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
% such that points within each community lie on \(K\)-dimensional
% orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
% removes the orthogonality property but retains the property that each
% community is represented by a \(K\)-dimensional subspace. Therefore, the
% ASE of \(P\) still consists of subspaces 
% that correspond to communities of the PABM, even if they are not orthogonal, 
% suggesting the use of SSC. Before exploring SSC, we will first consider
% a different approach.
\begin{theorem}
\label{theorem3}
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix. Let $B = n V V^\top$. Then $B_{ij} = 0$ if vertices $i$ and $j$ are
from different communities.
\end{theorem}
\begin{proof}
% \begin{lemma}
% \label{lemma1}
% Let $P = V D V^\top$ be the spectral decomposition of the edge probability 
% matrix for a PABM.
We first show that $V V^\top =
\tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top} $ where $X$ is 
defined as in Eq.~\eqref{eq:xy}. Indeed, by Theorem 2, 
\(P = \tilde{\Pi} X U I_{p, q} U^\top X^\top \tilde{\Pi}\) for $p = K(K+1)/2$ and $q = K(K-1)/2$. 
The eigendecomposition \(P = V D V^\top\) also yields $P = V
|D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ where \(|\cdot|^{1/2}\) is
applied entry-wise. Now let $Y = \tilde{\Pi} XU$ and $\tilde{Y} = V|D|^{1/2}$; note that
$Y$ and $\tilde{Y}$ both have full column ranks. 
Because $P = Y I_{p,q} Y^{\top} = \tilde{Y} I_{p,q} \tilde{Y}^{\top}$, we have
$$Y = \tilde{Y} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}.$$
Let $Q = I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}$ and note that
$Y = \tilde{Y} Q$. We then have
\begin{equation*}
  \begin{split}
  Q^{\top} I_{p,q} Q &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} \tilde{Y} I_{p,q}
I_{p,q} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} \\ &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} Y I_{p,q}
Y^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} =  I_{p,q}
\end{split}
\end{equation*}
and hence $Q$ is an indefinite orthogonal matrix.

@@

Under the condition that $\|\Lambda\|_F > 0$, 
Theorem \ref{theorem3} provides perfect community detection from \(P\).
\footnote{This is excluding the special case in which vectors in each 
subspace lie on mutually orthogonal sub-subspaces.}
More specifically let \(|B|\) be the affinity matrix for graph \(G'\), 
where $|\cdot|$ is applied entry-wise. Then
\(G'\) consists of exactly \(K\) disjoint subgraphs, 
as $G'$ has no edges between communities. 
All that is left to identify the communities is 
to assign each subgraph a distinct community label. 
In practice, we do not observe $P$ and instead only observe the noisy
$A \sim \BG(P)$. A natural approach is then to use
the affinity matrix $\hat{B} = n \hat{V} \hat{V}^{\top}$ where
$\hat{V}$ is the matrix of eigenvectors (corresponding to the largest
eigenvalues in modulus) of $A$. The resulting procedure, named
Orthogonal Spectral Clustering, is presented in
Algorithm~\ref{alg:osc}.
The following result leverages existing theoretical properties
of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
$\hat{B}$ converges almost surely to $B$; in particular 
 $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ 
for each pair $(i, j)$ in different communities.

@@

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
obtained from OSC as described in Algorithm~\ref{alg:osc}. Then for $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4a}
\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. In particular $\hat{B}_{ij} -
B_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$ where the convergence is
uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
communities we have
% \begin{equation} \label{eq:thm4}
% \max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
% O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
% \end{equation}
% with high probability. 
% Thus for all pairs $(i, j)$ in different communities,
% we have uniform convergence 
$\hat{B}_{ij} \overset{\mathrm{a.s}}{\rightarrow} 0$. 
\end{theorem}
Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of mis-clustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em
  proportion} of mis-clustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

\begin{theorem}
\label{theorem6}
Let each $\tilde{\lambda}^{(k \ell)}$ be the popularity vector derived
from its corresponding $P^{(k \ell)}$ and let $\hat{\lambda}^{(k
  \ell)}$ be its estimate obtained from $A^{(k \ell)}$ using Algorithm~\ref{alg:param_est}.
Then if $n \rho_n = \omega( \log^{4}{n})$,
\begin{equation} \label{eq:thm6}
\max_{k, \ell \in \{1, ..., K\}} 
\|\hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)}\|_{\infty} = 
O\bigg(\frac{\log n_k}{\sqrt{n_k}} \bigg)
\end{equation}
with high probability. Here $\|\cdot\|_{\infty}$ denotes the
$\ell_\infty$ norm of a vector. Let $\hat{\Lambda}$ be the matrix
$$\hat{\Lambda} = \begin{bmatrix} \hat{\lambda}^{(11)} &
  \hat{\lambda}^{(12)} & \cdots & \hat{\lambda}^{(1 K)} \\
  \hat{\lambda}^{(21)} &
  \hat{\lambda}^{(22)} & \cdots & \hat{\lambda}^{(2 K)} \\
  \vdots & \vdots & \cdots & \vdots \\
   \hat{\lambda}^{(K1)} &
  \hat{\lambda}^{(K2)} & \cdots & \hat{\lambda}^{(K K)} 
\end{bmatrix}$$
and let $\hat{P} = \hat{\Lambda} I_{p,q}
\hat{\Lambda}^{\top}$. Eq.~\eqref{eq:thm6} then implies
\begin{equation}
\label{eq:conv_prob}
\frac{1}{n} \|\rho_n^{-1} \hat{P} - \rho_n^{-1} P\|_{F} = O((n
\rho_n)^{-1/2}), \quad \max_{ij} |\rho_n^{-1} \hat{P}_{ij} -
\rho_n^{-1} P_{ij}| = O((n \rho_n)^{-1/2})
\end{equation}
with high probability. \end{theorem}
Eq.~\eqref{eq:thm6} guarantees that $n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} -
\Lambda\|_{F} = O((n \rho_n)^{-1/2})$. Eq.~\eqref{eq:conv_prob} then guarantees that
the mean square error for $\rho_n^{-1} (\hat{P} - P)$ converges to $0$ almost surely and
furthermore the entries of $\rho_n^{-1}
\hat{P}$ converge uniformly to the entries of $\rho_n^{-1} P$; recall that
$\rho_n^{-1} P_{ij} = \lambda_{iz_j} \lambda_{j}z_i$. We note that
these results are stronger than existing results in
\citet{307cbeb9b1be48299388437423d94bf1}; for example Theorem~2 in
\citet{307cbeb9b1be48299388437423d94bf1} only guarantees $n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} -
\Lambda\|_{F} = o(1)$ as $n \rightarrow \infty$.
\hypertarget{simulated-examples}{%
\section{Simulation Study}\label{simulated-examples}}

@@

\hypertarget{proofs}{%
\section{Proofs of Theorem~\ref{theorem4}, Theorem~\ref{theorem5}, and Theorem~\ref{theorem6}}}
Let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues (in modulus), respectively. 
We first state an important technical lemma for bounding the maximum
$\ell_2$ norm difference between the rows of $\hat{V}_n$ and
$V_n$. See \citet{cape_biometrika} and 
\citet[Lemma~5]{rubindelanchy2017statistical} for a proof. 
\begin{lemma}
  \label{lem:technical}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell \ell)}\}_{K})$ be a $K$-blocks
PABM graph on $n$ vertices and let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues in modulus, respectively.
Let \((v_n^{(i)})^\top\) and \((\hat{v}_n^{(i)})^\top\) denote the $i$th 
row of \(V_n\) and \(\hat{V}_n\), respectively. 
Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
that with high probability,
$$\max_{i} \|W \hat{v}_n^{(i)} - v_n^{(i)}\|  = O\Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
\end{lemma}

%
%then the \(ij\)\textsuperscript{th} element of
%\(n X (X^\top X)^{-1} X^\top = n V V^\top = B\) is 0.

\begin{proof}[Proof of Theorem~\ref{theorem4}]
  Recall the notations in Lemma~\ref{lem:technical} and note that,
  under our assumption that the latent vectors $\lambda^{(k \ell)}$
  are all homogeneous, we have $\max_{i} \|v_n^{(i)}\| =
  O(n^{-1/2})$. 
%   Let \(V_n\) and \(\hat{V}_n\)
% be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
% \(K (K + 1) / 2\) most positive and \(K (K - 1) / 2\) most negative
% eigenvalues, respectively. From Lemma~5 in
% \citet{rubindelanchy2017statistical} we have, for some
% \(W \in \mathbb{O}(K^2)\) and \(c > 0\), that
% $$\|\hat{V} W - V\|_{2 \to \infty} = O_P \big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \big).$$
% We furthermore have \(\|V\|_{2 \to \infty} = O_P(n^{-1/2})\). 
% Let
% \((v_n^{(i)})^\top\) and \((\hat{v}_n^{(i)})^\top\) denote the $i$th 
% row of \(V_n\) and \(\hat{V}_n\), respectively. 
Next recall Theorem~\ref{theorem3}; in particular $B_{ij} = n(v_n^{(i)})^{\top}
v_n^{(j)}$. % Now suppose that vertices $i$ and $j$ belongs to different
% communities. Then $B_{ij} = 0$ and hence
% Then for \(i\) and
% \(j\) belonging to different
% communities, we ha using the fact that \((v_n^{(i)})^\top v_n^{(j)} = 0\),
We therefore have
\[\begin{split}
 \max_{ij} |\hat{B}_{ij} - B_{ij}| &= \max_{ij} n |(\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)} -
(v_n^{(i)})^\top v_n^{(j)}| \\
& \leq n \max_{ij} |(\hat{v}_n^{(i)})^\top W W^\top \hat{v}_n^{(j)} -
(v_n^{(i)})^\top v_n^{(j)}| \\
& \leq n \max_{i,j} \Bigl(\|W^{\top} \hat{v}_n^{(i)} - v_n^{(i)}\| \times \|\hat{v}_n^{(j)}\|
+ \|W^{\top} \hat{v}_n^{(j)} - v_n^{(j)}\| \times \|v_n^{(i)}\|\Bigr) \\
% & \leq 
% \hat{V}_n W^\top W \hat{V}_n - V_n V_n^\top||_{2 \to \infty} \\
% & = ||2 \hat{V}_n W V_n^\top - 2 V_n V_n^\top + \hat{V}_n W W^\top V_n^\top - 
% 2 \hat{V}_n W V_n^\top + V_n V_n^\top||_{2 \to \infty} \\
% & = ||2 (\hat{V}_n W - V_n) V_n^\top + 
% (\hat{V}_n W - V_n) (\hat{V}_n W - V_n)^\top ||_{2 \to \infty} \\
% & \leq 2 
% &\leq  n \Bigl(\max_{ij}  \|W \hat{v}_n^{(i)}  - v_n^{(i)} \| \times \|\hat{v}_n^{(j)}\| +
% n \|W \hat{v}_n^{(j)}  - v_n^{(j)} \| \times \|v_n^{(i)}\|\Bigr)
% \\
& \leq n \Bigl(\max_{ij}  \|W \hat{v}_n^{(i)}  - v_n^{(i)} \|^2 +  \|W
\hat{v}_n^{(i)}  - v_n^{(i)} \| \times \|v_n^{(j)}\| +  \|W \hat{v}_n^{(j)}  - v_n^{(j)} \| \times \|v_n^{(i)}\|\Bigr) 
\\ &
\leq n \max_{i} \|W \hat{v}_n^{(i)}  - v_n^{(i)} \|^2 + 2n \max_{i}
\|W \hat{v}_n^{(i)}  - v_n^{(i)} \| \times \max_{j} \|v_n^{(j)}\|
\\
& = O \Big( \frac{\log^{c}{n}}{n^{1/2} \rho_n^{1/2}} \Big)
\end{split}\]
with high probability.
Theorem~\ref{theorem4} follows from the above bound together with the
conclusion in Theorem~\ref{theorem3} that $B_{ij} = 0$ whenever vertices $i$ and $j$
belongs to different communities. 
\end{proof}
% Scaling all terms in the above inequality by \(n\), we obtain
% $$|n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)$$
% as desired.

>>>>

\author{John Koo\\Department of Statistics, Indiana University\\ \\
Minh Tang\\Department of Statistics, North Carolina State University\\ \\
Michael W. Trosset\\Department of Statistics, Indiana University\\ \\
September 8, 2021}

@@

Let $G = (V, E)$ be an unweighted, undirected, 
and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. The matrix
$A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ 
such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ 
and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. 
We further restrict our analyses to Bernoulli graphs. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. 
Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ 
for each $1 \leq i < j \leq n$ (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). 
We denote $A \sim \BG(P)$ as a graph with adjacency matrix $A$ 
sampled from edge probability matrix $P$ in this manner. 
If each vertex has a hidden label in $[K]$, 
they are denoted as $z_1, ..., z_n$. 
$\lambda_{ik}$ denotes the popularity parameter of vertex $i$ to community $k$. 
$\Lambda$ is the $n \times K$ matrix of popularity parameters. 
Finally, we denote 
$X = \bigl[x_1 \mid \cdots \mid x_n\bigr]^\top \in \mathbb{R}^{n \times d}$ 
as the matrix corresponding to a collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

@@

Nevertheless by using the special geometric structure of $X$ we can circumvent the
non-identifiability of $Y$ and $XU$ by using instead the rows of the
matrix $V$ of eigenvectors (corresponding to the non-zero eigenvalues) of $P$. In particular $V$ is identifiable
up to orthogonal transformations and furthermore, due to the block
diagonal structure of $X$, the rows of $V$ also lie on $K$ distinct orthogonal
subspaces and hence $v_i^{\top} v_j = 0$ whenever $z_i \not = z_j$. 
% The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
% such that points within each community lie on \(K\)-dimensional
% orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
% removes the orthogonality property but retains the property that each
% community is represented by a \(K\)-dimensional subspace. Therefore, the
% ASE of \(P\) still consists of subspaces 
% that correspond to communities of the PABM, even if they are not orthogonal, 
% suggesting the use of SSC. Before exploring SSC, we will first consider
% a different approach.
\begin{theorem}
\label{theorem3}
Let $P = V D V^\top$ be the spectral decomposition 
of the edge probability matrix. 
Let $B = n V V^\top$. 
Assume $\lambda_{i z_i} > 0$ for each $i \in [n]$, 
i.e., each vertex's popularity parameter to its own community is nonzero. 
Then if vertices $i$ and $j$ are from different communities, $B_{ij} = 0$.
\end{theorem}
\begin{proof}
% \begin{lemma}
% \label{lemma1}
% Let $P = V D V^\top$ be the spectral decomposition of the edge probability 
% matrix for a PABM.
We first show that $V V^\top =
\tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top} $ where $X$ is 
defined as in Eq.~\eqref{eq:xy}. Indeed, by Theorem 2, 
\(P = \tilde{\Pi} X U I_{p, q} U^\top X^\top \tilde{\Pi}\) for $p = K(K+1)/2$ and $q = K(K-1)/2$. 
The eigendecomposition \(P = V D V^\top\) also yields $P = V
|D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ where \(|\cdot|^{1/2}\) is
applied entry-wise. Now let $Y = \tilde{\Pi} XU$ and $\tilde{Y} = V|D|^{1/2}$; note that
$Y$ and $\tilde{Y}$ both have full column ranks. 
Because $P = Y I_{p,q} Y^{\top} = \tilde{Y} I_{p,q} \tilde{Y}^{\top}$, we have
$$Y = \tilde{Y} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}.$$
Let $Q = I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}$ and note that
$Y = \tilde{Y} Q$. We then have
\begin{equation*}
  \begin{split}
  Q^{\top} I_{p,q} Q &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} \tilde{Y} I_{p,q}
I_{p,q} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} \\ &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} Y I_{p,q}
Y^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} =  I_{p,q}
\end{split}
\end{equation*}
and hence $Q$ is an indefinite orthogonal matrix.

@@

Theorem \ref{theorem3} provides perfect community detection from \(P\)\footnote{This excludes the pathological case in which $P_{ij} = 0$ for some $z_i = z_j$.}.
More specifically, let \(|B|\) be the affinity matrix for graph \(G'\), 
where $|\cdot|$ is applied entry-wise. Then
\(G'\) consists of exactly \(K\) disjoint subgraphs, 
as $G'$ has no edges between communities. 
All that is left to identify the communities is 
to assign each subgraph a distinct community label. 
In practice, we do not observe $P$ and instead only observe the noisy
$A \sim \BG(P)$. A natural approach is then to use
the affinity matrix $\hat{B} = n \hat{V} \hat{V}^{\top}$ where
$\hat{V}$ is the matrix of eigenvectors (corresponding to the largest
eigenvalues in modulus) of $A$. The resulting procedure, named
Orthogonal Spectral Clustering, is presented in
Algorithm~\ref{alg:osc}.
The following result leverages existing theoretical properties
of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
$\hat{B}$ converges almost surely to $B$; in particular 
 $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ 
for each pair $(i, j)$ in different communities.

@@

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
obtained from OSC as described in Algorithm~\ref{alg:osc}. Then for $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4a}
\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. In particular $\hat{B}_{ij} -
B_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$ where the convergence is
uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
communities we have
% \begin{equation} \label{eq:thm4}
% \max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
% O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
% \end{equation}
% with high probability. 
% Thus for all pairs $(i, j)$ in different communities,
% we have uniform convergence 
$\hat{B}_{ij} \overset{\mathrm{a.s}}{\rightarrow} 0$. 
\end{theorem}
Theorem~\ref{theorem4} guarantees that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability
converging to one as $n$ increases. 
Furthermore, if $P_{ij} > 0$ for each pair $(i, j)$ in the same community, 
we can always find an $\epsilon > 0$ such that $\hat{B}_{ij} > \epsilon$ 
with probability converging to one as $n$ increases. 
Thus, by using $\hat{B}$, we can 
perfectly recover all the latent community assignments $z_1, z_2,
\dots, z_n$, i.e., the number of mis-clustered vertices is zero
asymptotically almost surely. We note that Theorem~\ref{theorem4} is
stronger than existing results in the literature; in particular
Theorem~1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
originally introduces the PABM model) only guarantees that the {\em
  proportion} of mis-clustered vertices converges to $0$ as $n
\rightarrow \infty$. Furthermore Theorem~1
of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity
parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which
is a considerably stronger assumption than the assumption $n \rho_n =
\omega(\log^{4}{n})$ used in Theorem~\ref{theorem4}. Indeed, $n
\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for
some constant $c > 1$ is commonly used in the context of graph
estimation using spectral methods.

@@

\begin{theorem}
\label{theorem6}
Let each $\tilde{\lambda}^{(k \ell)}$ be the popularity vector derived
from its corresponding $P^{(k \ell)}$ and let $\hat{\lambda}^{(k
  \ell)}$ be its estimate obtained from $A^{(k \ell)}$ using Algorithm~\ref{alg:param_est}.
Then if $n \rho_n = \omega( \log^{4}{n})$,
\begin{equation} \label{eq:thm6}
\max_{k, \ell \in \{1, ..., K\}} 
\|\hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)}\|_{\infty} = 
O\bigg(\frac{\log n_k}{\sqrt{n_k}} \bigg)
\end{equation}
with high probability. Here $\|\cdot\|_{\infty}$ denotes the
$\ell_\infty$ norm of a vector. Let $\hat{\Lambda}$ be the matrix

$$\hat{\Lambda} = \begin{bmatrix} \hat{\lambda}^{(11)} &
  \hat{\lambda}^{(12)} & \cdots & \hat{\lambda}^{(1 K)} \\
  \hat{\lambda}^{(21)} &
  \hat{\lambda}^{(22)} & \cdots & \hat{\lambda}^{(2 K)} \\
  \vdots & \vdots & \cdots & \vdots \\
   \hat{\lambda}^{(K1)} &
  \hat{\lambda}^{(K2)} & \cdots & \hat{\lambda}^{(K K)} 
\end{bmatrix}$$

and let $\hat{P} = \hat{X} U I_{p,q} U^\top \hat{X}^{\top}$ 
where $\hat{X}$ is defined from $\hat{\Lambda}$ and $U$ is defined from $K$ as in Theorem~\ref{theorem2}. 
Eq.~\eqref{eq:thm6} then implies
\begin{equation}
\label{eq:conv_prob}
\frac{1}{n} \|\rho_n^{-1} \hat{P} - \rho_n^{-1} P\|_{F} = O((n
\rho_n)^{-1/2}), \quad \max_{ij} |\rho_n^{-1} \hat{P}_{ij} -
\rho_n^{-1} P_{ij}| = O((n \rho_n)^{-1/2})
\end{equation}
with high probability. \end{theorem}
Eq.~\eqref{eq:thm6} guarantees that $n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} -
\Lambda\|_{F} = O((n \rho_n)^{-1/2})$. Eq.~\eqref{eq:conv_prob} then guarantees that
the mean square error for $\rho_n^{-1} (\hat{P} - P)$ converges to $0$ almost surely and
furthermore the entries of $\rho_n^{-1}
\hat{P}$ converge uniformly to the entries of $\rho_n^{-1} P$; recall that
$\rho_n^{-1} P_{ij} = \lambda_{iz_j} \lambda_{j z_i}$. We note that
these results are stronger than existing results in
\citet{307cbeb9b1be48299388437423d94bf1}; for example Theorem~2 in
\citet{307cbeb9b1be48299388437423d94bf1} only guarantees $n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} -
\Lambda\|_{F} = o(1)$ as $n \rightarrow \infty$.

\hypertarget{simulated-examples}{%
\section{Simulation Study}\label{simulated-examples}}

@@

\hypertarget{proofs}{%
\section{Proofs of Theorem~\ref{theorem4}, Theorem~\ref{theorem5}, and Theorem~\ref{theorem6}}}
Let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues (in modulus), respectively. 
We first state an important technical lemma for bounding the maximum
$\ell_2$ norm difference between the rows of $\hat{V}_n$ and
$V_n$. See \citet{cape_biometrika} and 
\citet[Lemma~5]{rubindelanchy2017statistical} for a proof.

\begin{lemma}
\label{lem:technical}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell \ell)}\}_{K})$ be a $K$-blocks
PABM graph on $n$ vertices and let \(V\) and \(\hat{V}\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of 
\(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues in modulus, respectively.
Let \(v_i^\top\) and \(\hat{v}_i^\top\) denote the $i$th 
row of \(V\) and \(\hat{V}\), respectively. 
Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
that with high probability,
$$\max_{i} \|W \hat{v}_i - v_i\|  = O\Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
\end{lemma}

\begin{proof}[Proof of Theorem~\ref{theorem4}]
Recall the notations in Lemma~\ref{lem:technical} and note that,
under our assumption that the latent vectors $\lambda^{(k \ell)}$
are all homogeneous, we have $\max_{i} \|v_i\| =
O(n^{-1/2})$.

Next recall Theorem~\ref{theorem3}; in particular $B_{ij} = nv_i^{\top}
v_j$. 
We therefore have
\[\begin{split}
 \max_{ij} |\hat{B}_{ij} - B_{ij}| &= \max_{ij} n |\hat{v}_i^\top \hat{v}_j -
v_i^\top v_j| \\
& \leq n \max_{ij} |\hat{v}_i^\top W W^\top \hat{v}_j -
v_i^\top v_j| \\
& \leq n \max_{i,j} \Bigl(\|W^{\top} \hat{v}_i - v_i\| \times \|\hat{v}_j\|
+ \|W^{\top} \hat{v}_j - v_j\| \times \|v_i\|\Bigr) \\
& \leq n \Bigl(\max_{ij}  \|W \hat{v}_i  - v_i \|^2 +  \|W
\hat{v}_i  - v_i \| \times \|v_j\| +  \|W \hat{v}_j  - v_j \| \times \|v_i\|\Bigr) 
\\ &
\leq n \max_{i} \|W \hat{v}_i  - v_i \|^2 + 2n \max_{i}
\|W \hat{v}_i  - v_i \| \times \max_{j} \|v_j\|
\\
& = O \Big( \frac{\log^{c}{n}}{n^{1/2} \rho_n^{1/2}} \Big)
\end{split}\]
with high probability.
Theorem~\ref{theorem4} follows from the above bound together with the
conclusion in Theorem~\ref{theorem3} that $B_{ij} = 0$ whenever vertices $i$ and $j$
belongs to different communities. 
\end{proof}

We now provide a proof of Theorem~\ref{theorem5}. Our proof is based
on verifying the sufficient conditions given in Theorem~6 of
\citet{jmlr-v28-wang13}
under which sparse subspace clustering based on solving the
optimization problem in Eq.~\eqref{eq:ssc} yields an affinity matrix
$B = |C| + |C^{\top}|$ satisfying the subspace detection property of
Definition~\ref{def:subspace_detection}. We first recall a few
definitions used in \citet{soltanolkotabi2012} and \citet{jmlr-v28-wang13}; for ease of exposition,
these definitions are stated using the notations of the current
paper and we will drop the explicit dependency on $n$ from our
eigenvectors $\hat{V}$ of $A$ and $V$ of $P$.
 % We first recall the
% notion of the inradius of a set of points. 
\begin{definition}[Inradius]
  \label{inradius}
The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is
defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
Let $X$ be a $n \times d$ matrix with rows $x_1, x_2, \dots,
x_n$. We then define, with a slight abuse of notation, $r(X)$ as the
inradius of the convex hull formed by $\{\pm x_1, \pm x_2, \dots, \pm x_n\}$. 
\end{definition}
